### 2012-03-20 doc/theses/mikula/proposed_algorithm.tex
Master thesis: In progress.

Additionally, PathType contains a list of special functions that were called with an argument of PathType type. The motivation is that in some cases of function calls, we want to know that the function call is performed with an instance of PathType because then, we can determine a type of the function call more precisely. Those special functions are built-in functions \texttt{data}, \texttt{min}, \texttt{max}, \texttt{avg}, \texttt{sum}, \texttt{distinct-values}. \todo{DOPLNIT}

@@

\begin{figure}
\label{FIG_syntax_tree_1_types}
\caption{Sample syntax tree 1 after the static analysis of types.}
\includegraphics[scale=0.6]{syntax_tree_1_types.eps}
\end{figure}

@@

\subsection{Function Calls}
This case is quite straightforward. The algorithm encounters a function call and one of the arguments is a set of elements of attributes (subexpression $E$ representing $S$) represented by PathType $P$. To determine the type of $S$, it is only needed to determine the type of the corresponding formal argument from the definition of the function. The function is either a built-in one so its definition is known or it is defined in the prolog section. External functions are not processed as was mentioned already.

If the type $T$ of the formal argument is a built-in type or its sequence, then $T$ is also the inferred type of $S$. The inferred statement is $P \rightarrow T$. Otherwise, no statement is inferred.

\subsection{Arithmetic Operations}
If the operator in an arithmetic operation is one of \texttt{+}, \texttt{-}, \texttt{div}, \texttt{mod}, \texttt{*}, \texttt{/} (the class of the expression node is \texttt{Operator} and it represents one of \texttt{PLUS}, \texttt{MINUS}, \texttt{IDIV}, \texttt{MOD}, \texttt{MUL}, \texttt{DIV} (for all operators, see Attachment \ref{FIG_operators})), one operand is of PathType $P$ and the type $T$ of the other operand is one of numeric built-in types, then the inferred statement is $P \rightarrow T$.

If the operator is one of \texttt{<}, \texttt{>}, \texttt{<=}, \texttt{>=}, \texttt{=}, \texttt{!=} (the class of the expression node is \texttt{Operator} and it represents one of \texttt{GEN\_LESS\_THAN}, \texttt{GEN\_GREATER\_THAN}, \texttt{GEN\_LESS\_THAN\_EQUALS}, \texttt{GEN\_GREATER\_THAN\_EQUALS}, \texttt{GEN\_EQUALS}, \\ \texttt{GEN\_NOT\_EQUALS}), one operand is of PathType $P$ and the type $T$ of the other operand is one of built-in types, then the inferred statement is $P \rightarrow T$.

@@

\todo[inline]{Pěkné, neškodilo by proložit příklady, tj. dotaz, strom, co z něho odvodíme. -- To urobim pocas implementacie pre nejaky vzorovy dotaz.}

>>>>

Additionally, PathType contains a list of special functions that were called with an argument of PathType type. The motivation is that in some cases of function calls, we want to know that the function call is performed with an instance of PathType because then, we can determine a type of the function call more precisely. Those special functions are built-in functions \texttt{data}, \texttt{min}, \texttt{max}, \texttt{avg}, \texttt{sum}, \texttt{distinct-values}, \texttt{zero-or-one}, \texttt{exactly-one}. And other may be added, when needed.

@@

\begin{figure}
\label{FIG_syntax_tree_1_types}
\caption{Sample syntax tree 1 after the static analysis of types.}
\includegraphics[scale=0.5]{syntax_tree_1_types.eps}
\end{figure}

@@

\subsection{Function Calls}\label{SUBSECTION_function_calls}
This case is quite straightforward. The algorithm encounters a function call and one of the arguments is a set of elements of attributes (subexpression $E$ representing $S$) represented by PathType $p$. To determine the type of $S$, it is only needed to determine the type of the corresponding formal argument from the definition of the function. The function is either a built-in one so its definition is known or it is defined in the prolog section. External functions are not processed as was mentioned already.

If the type $t$ of the formal argument is a built-in type or its sequence, then $t$ is also the inferred type of $S$. The inferred statement is $p \rightarrow t$. Otherwise, no statement is inferred.

\subsection{Arithmetic Operations}
If the operator in an arithmetic operation is one of \texttt{+}, \texttt{-}, \texttt{div}, \texttt{mod}, \texttt{*}, \texttt{/} (the class of the expression node is \texttt{Operator} and it represents one of \texttt{PLUS}, \texttt{MINUS}, \texttt{IDIV}, \texttt{MOD}, \texttt{MUL}, \texttt{DIV} (for all operators, see Attachment \ref{FIG_operators})), one operand is of PathType $p$ and the type $t$ of the other operand is one of numeric built-in types, then the inferred statement is $p \rightarrow t$.

If the operator is one of \texttt{<}, \texttt{>}, \texttt{<=}, \texttt{>=}, \texttt{=}, \texttt{!=} (the class of the expression node is \texttt{Operator} and it represents one of \texttt{GEN\_LESS\_THAN}, \texttt{GEN\_GREATER\_THAN}, \texttt{GEN\_LESS\_THAN\_EQUALS}, \texttt{GEN\_GREATER\_THAN\_EQUALS}, \texttt{GEN\_EQUALS}, \\ \texttt{GEN\_NOT\_EQUALS}), one operand is of PathType $p$ and the type $t$ of the other operand is one of built-in types, then the inferred statement is $p \rightarrow t$.

\subsection{Example}

\begin{figure}
\label{FIG_syntax_tree_1_types_type_inference}
\caption{Example of an inference of a type from a function call.}
\includegraphics[scale=1]{syntax_tree_1_types_type_inference.eps}
\end{figure}

Figure \ref{FIG_syntax_tree_1_types_type_inference} shows a fraction of the syntax tree from Figure \ref{FIG_syntax_tree_1_types}. When the inference of built-in types encounters the node marked by blue color, a statement will be inferred using the principle described in Chapter \ref{SUBSECTION_function_calls}. The node represents a call of \texttt{local:convert} function.

This function has one formal argument of type $t =$ \texttt{xs:decimal}, \texttt{zero-or-one} cardinality. The real argument is of a PathType $p$. In particular, it is a PathType representing path \texttt{\$i/reserve}, where the \texttt{\$i} variable refers to a PathType representing for-bound \texttt{/site/open\_auctions/open\_auction} path, and one special function call of \texttt{zero-or-one} function is noted.

Since the criteria for the inference from a function call are met, statement $p \rightarrow t$ is inferred.

@@
